---
title: "Intro to R, Assignment 1"
geometry: margin=0.75in
output:
  pdf_document: default
  word_document: default
  html_document: default
---


# A brief introduction to R packages  
Packages are curated collections of functions and/or compiled code for use in R. Some commonly-downloaded, popular packages include sample data and documentation to explain and demonstrate implementation.  
Packages can be downloaded from the Comprehensive R Archive Network (CRAN), the primary repository for R packages. Each package submitted to CRAN is reviewed and tested thoroughly before release. Every package is released with a version number, and the most popular packages are frequently updated and released with a new version number.  
CRAN is a trusted source for up-to-date, consistent versions of R packages, and allows package accessibility across multiple servers in multiple regions called 'mirrors'. Packages can also be downloaded through GitHub, but we'll get to that later.  


# 1. Using `install.packages()`  
One of the most useful packages in R is `tidyverse`.  
Install `tidyverse` by typing `install.packages('tidyverse')` in the code chunk below.  
The `install.packages()` function is the easiest method to download packages from CRAN.  
View the documentation for `install.packages()` by clicking [**here.**](https://www.rdocumentation.org/packages/utils/versions/3.6.2/topics/install.packages)  

```{r}



```


# 2. The `library()` function  
Once the package is downloaded, the package must be called into virtual memory before it is used. In order to use the package, call it into virtual memory with the `library()` command.  
When downloading packages, the package name must be in quotation marks. When called into virtual memory with the `library()` command, no quotation marks are needed.  

Put the `tidyverse` package into virtual memory by typing `library(tidyverse)` in the code chunk below.  

```{r}



```


# 3. Viewing package version with `packageVersion()`  
In an R Markdown document, you can begin a new code chunk by adding three tick marks followed by a lower case r enclosed in curly brackets: ` ```{r} `  

You can close the code chunk with three tick marks: ` ``` `.  

Create a code chunk below and find the package version of tidyverse you've just downloaded by using the `packageVersion()` function.  
View the documentation for `packageVersion()` by clicking [**here.**](https://www.rdocumentation.org/packages/Biobase/versions/2.32.0/topics/package.version)  

```{r}



```


# 4. Importing a .csv file with `read.csv`  
Read the .csv file titled `brfss2024tri.csv` into a dataframe with the `read.csv()` function. Name the resulting dataframe `brtri`.  
Hint: include the file path when reading the file, and remember to either double backslashes (`\`) or use forward slashes (`/`) in the file path.  
View the documentation for `read.csv()` by clicking [**here.**](https://www.rdocumentation.org/packages/COVID19/versions/2.0.3/topics/read.csv)  

```{r}



```


# 5. The `colnames()` function  
View the column names in the `brtri` dataframe with the `colnames()` function in the code chunk below.  
How many columns are in the dataframe?  
View the documentation for `colnames()` by clicking [**here.**](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/row+colnames)  
View additional methods for retrieving column names and details by clicking [**here.**](https://www.r-bloggers.com/2024/10/mastering-column-names-in-base-r-a-beginners-guide/)  

```{r}



```


# 6. Using `sapply()` to view column type  
View the class of each column in the `brtri` dataframe by implementing the `sapply()` function with the `class()` function in the code chunk below.  
View the documentation for `sapply()` by clicking [**here.**](https://www.rdocumentation.org/packages/functools/versions/0.2.0/topics/Sapply)  

```{r}



```


# 7. The `table()` function  
Use the `table()` function to view the frequencies of the `STATE` column in the `brtri` dataframe in the code chunk below. Use the parameter `useNA` to view the number of rows with missing data.  
View the documentation for `table()` by clicking [**here.**](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/table)  

```{r}



```


# 8. Viewing the number of rows with `nrow()`  
Use the `nrow()` function to view the number of rows in the `brtri` dataframe in the code chunk below.  
View the documentation for `nrow()` by clicking [**here.**](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/nrow)  

```{r}



```



