---
title: "Intro to R, Assignment 2"
geometry: margin=0.75in
output:
  pdf_document: default
  word_document: default
  html_document: default
---


# A brief introduction to R packages  
Packages are curated collections of functions and/or compiled code for use in R. Some commonly-downloaded, popular packages include sample data and documentation to explain and demonstrate implementation.  
Packages can be downloaded from the Comprehensive R Archive Network (CRAN), the primary repository for R packages. Each package submitted to CRAN is reviewed and tested thoroughly before release. Every package is released with a version number, and the most popular packages are frequently updated and released with a new version number.  
CRAN is a trusted source for up-to-date, consistent versions of R packages, and allows package accessibility across multiple servers in multiple regions called 'mirrors'. Packages can also be downloaded through GitHub, but we'll get to that later.  


# 1. Using `install.packages()`  
One of the most useful packages in R is `tidyverse`.  
Install `tidyverse` by typing `install.packages('tidyverse')` in the code chunk below.  
The `install.packages()` function is the easiest method to download packages from CRAN.  
View the documentation for `install.packages()` by clicking [**here.**](https://www.rdocumentation.org/packages/utils/versions/3.6.2/topics/install.packages)  

```{r eval=FALSE}

install.packages('tidyverse')

```


# 2. The `library()` function  
Once the package is downloaded, the package must be called into virtual memory before it is used. In order to use the package, call it into virtual memory with the `library()` command.  
When downloading packages, the package name must be in quotation marks. When called into virtual memory with the `library()` command, no quotation marks are needed.  

Put the `tidyverse` package into virtual memory by typing `library(tidyverse)` in the code chunk below.  

```{r message=FALSE, warning=FALSE}

library(tidyverse)

```


# 3. Viewing package version with `packageVersion()`  
In an R Markdown document, you can begin a new code chunk by adding three tick marks followed by a lower case r enclosed in curly brackets: ` ```{r} `  

You can close the code chunk with three tick marks: ` ``` `.  

Create a code chunk below and find the package version of tidyverse you've just downloaded by using the `packageVersion()` function.  
View the documentation for `packageVersion()` by clicking [**here.**](https://www.rdocumentation.org/packages/Biobase/versions/2.32.0/topics/package.version)  

```{r}

packageVersion('tidyverse')

```


# 4. Importing a .csv file with `read.csv`  
Read the .csv file titled `brfss2024tri.csv` into a dataframe with the `read.csv()` function. Name the resulting dataframe `brtri`.  
Hint: include the file path when reading the file, and remember to either double backslashes (`\`) or use forward slashes (`/`) in the file path.  
View the documentation for `read.csv()` by clicking [**here.**](https://www.rdocumentation.org/packages/COVID19/versions/2.0.3/topics/read.csv)  

```{r}

brtri <- read.csv("brfss2024tri.csv")

```


# 5. The `colnames()` function  
View the column names in the `brtri` dataframe with the `colnames()` function in the code chunk below.  
How many columns are in the dataframe?  
View the documentation for `colnames()` by clicking [**here.**](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/row+colnames)  
View additional methods for retrieving column names and details by clicking [**here.**](https://www.r-bloggers.com/2024/10/mastering-column-names-in-base-r-a-beginners-guide/)  

```{r}

colnames(brtri)

```


# 6. Using `sapply()` to view column type  
View the class of each column in the `brtri` dataframe by implementing the `sapply()` function with the `class()` function in the code chunk below.  
View the documentation for `sapply()` by clicking [**here.**](https://www.rdocumentation.org/packages/functools/versions/0.2.0/topics/Sapply)  

```{r}

sapply(brtri, class)

```


# 7. The `table()` function  
Use the `table()` function to view the frequencies of the `STATE` column in the `brtri` dataframe in the code chunk below. Use the parameter `useNA` to view the number of rows with missing data.  
View the documentation for `table()` by clicking [**here.**](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/table)  

```{r warning=FALSE, message=FALSE}

table(brtri[,"STATE"], useNA = "ifany")

```


# 8. Viewing the number of rows with `nrow()`  
Use the `nrow()` function to view the number of rows in the `brtri` dataframe in the code chunk below.  
View the documentation for `nrow()` by clicking [**here.**](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/nrow)  

```{r}

nrow(brtri)

```


# 9. The BRFSS Codebook: Height  
The Behavioral Risk Factor Surveillance Survey (BRFSS) data in the `brtri` dataframe have a codebook with the following information:  

Label: Reported Height in Feet and Inches  
Section Name: Demographics  

Type of Variable: Num  
Variable Name: HEIGHT3  

Question:  About how tall are you without shoes?  (If respondent answers in metrics, put a 9 in the first column)[Round fractions down.]  

| Value | Value Label | Frequency | Percentage | Weighted Percentage |  
|:------------|:---------------------------------|----------:|-----------:|--------------------:|  
| 200-711 | Height (ft/inches) | | | |  
| | Notes: 0 _ / _ _ = feet / inches | 425,837 | 95.85 | 94.10 |    
| 7777 | Don't know/Not sure | 4,198 | 0.94 | 1.49 |  
| 9061-9998 | Height (meters/centimeters) | | | |  
| | Notes: The initial '9' indicates | | | |  
| | this was a metric value. | 7,793 | 1.75 | 2.91 |    
| 9999 | Refused | 6,442 |        1.45 | 1.50 |  
| BLANK | Not asked or Missing | 13,400 | | |  
  
  
Use the `table()` function to get a frequency table of the data in the `HEIGHT3` column in the code chunk below.  
Given the information in the table above, what range of values would you say are in feet and inches in column `HEIGHT3`?  
What range of values would you say are in centimeters?  
View the documentation for `table()` by clicking [**here.**](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/table)  

```{r}



```


# 10. Extracting height: feet, inches  
According to the information in the BRFSS codebook, values from 200 to 711 in the `HEIGHT3` column represent height in feet and inches.  
For example, the value '511' actually represents a height of 5 feet and 11 inches.  
The codebook also explains that values beginning with the number '9' in the `HEIGHT3` column precede digits that represent height recorded in centimeters.  
The number '9' only signifies that the remaining digits are recorded in centimeters and should not be included in the measurement values.  
In order to compare height of each respondent, we will need to convert all height values to inches or centimeters.  

We will accomplish this step by step:
1. Where height is recorded in feet and inches, for example '511', we'11 separate values of feet and values of inches into separate `HEIGHTft` and `HEIGHTin` columns.
2. Where height is measured in centimeters, for example '9170', we will need to extract the last three digits--the values following the leading'9'--and store as values of centimeters in a separate `HEIGHTcm` column.
3. After we have three new columns (`HEIGHTft`, `HEIGHTin`, `HEIGHTcm`), we can convert values conditionally by row until each respondent has height recorded in inches and centimeters.  

Extract the 'foot' (the first digit) and 'inch' (the last two digits) components of each respondent's height from the `HEIGHT3` column conditionally--only if `brtri$HEIGHT3>=200 & brtri$HEIGHT3<=711`.  
Use conditional methods with logical indexing and the `substr()` function to create `brtri$HEIGHTft` and `brtri$HEIGHTin` columns and populate with the correct digits from the `HEIGHT3` column.  
When subsetting columns, `NA` values are not allowed, so remember to include `!is.na(brtri$HEIGHT3)` in the logical indexing.  

You can use Stack Overflow as a resource on conditional methods with logical indexing by clicking [**here.**](https://stackoverflow.com/questions/13871614/replacing-values-from-a-column-using-a-condition-in-r)  
View the documentation for `substr()` by clicking [**here.**](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/substr)  

A brief note on a technique in data cleaning known as 'conditional' or 'logical indexing':  


In R, an **index** refers to the position used to locate a specific piece of data within a data structure—for a dataframe, this means specifying which row and which column you want to access, since data is organized in a two-dimensional grid.  
In code, this indexing is expressed using square brackets in the form `dataframe_name[row, column]`.  
For example, `dataframe_name[2, 3]` accesses the value in the second row and third column, while `dataframe_name[1:5, c('column1', 'column2')]` selects rows 1 through 5 for two named columns.  
Importantly, if you omit the row or column specification, R interprets this as "all rows" or "all columns", so `dataframe_name[, 'column1']` returns every row for `column1`, and `dataframe_name[3, ]` returns every column for row 3.  


**Logical indexing** extends this concept by using logical (TRUE/FALSE) conditions in place of explicit row or column numbers.  
Instead of specifying `dataframe_name[2, ]`, you might write `dataframe_name[dataframe_name$age > 30, ]`, which evaluates the condition for every row and returns only those where it's `TRUE`.  
This is especially powerful for conditional assignment: for instance, `dataframe_name$status[dataframe_name$age > 30] <- "adult"` will assign the value `"adult"` only to the rows where the age condition holds true, leaving all other rows untouched.  
This lets you selectively modify subsets of data without writing loops or manually identifying row numbers.  


```{r}



```


# 11. Transforming height: total inches  
Create a new column called `HEIGHTtotin` and populate it with values in total inches using column operations and the new `HEIGHTft` and `HEIGHTin` columns we just created.  
You should be able to create the column and populated it with values in a single line of code.  

```{r}



```


# 12. Extracting height: centimeters  
Create a new column called `HEIGHTcm` and populate it with `NA` values.  
Extract the height of each respondent recorded in centimeters from the `HEIGHT3` column conditionally--only if `brtri$HEIGHT3>=9061 & brtri$HEIGHT3<=9998`.  
Use conditional methods with logical indexing and the `substr()` function to populate with the correct digits from the `HEIGHT3` column.  
Remember that the last three digits of the `HEIGHT3` column represent respondent height in centimeters if the leading digit is a '9'.  
When subsetting columns, `NA` values are not allowed, so remember to include `!is.na(brtri$HEIGHT3)` in the logical indexing.  

```{r}



```


# 13. Transforming height: total centimeters  
Create a new column called `HEIGHTtotcm` by simply copying the column `brtri$HEIGHTcm`.  
This new column `HEIGHTtotcm` will hold the height of each respondent in centimeters.  
For rows with height recorded in inches, we will convert from inches to centimeters and populate `HEIGHTtotcm` with converted values.  

Use the `round()` function to round to the nearest whole centimeter.  

View the documentation for `round()` by clicking [**here.**](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/Round)  

```{r}



```


# 14. Transforming height: total meters  
Create a new column called `HEIGHTtotm` by simply converting the column `brtri$HEIGHTtotcm` to meters.  
Use the `round()` function to round values to four decimal places.  

```{r}



```


# 15. Transforming height: total centimeters to total inches  
Create a new column called `HEIGHTtotin` by simply converting the column `brtri$HEIGHTtotcm` to inches.  
Use the `round()` function to round values to the nearest whole inch.  

```{r}



```